---
output: 
  pdf_document:
    
    
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "Temporal, Spatial, and Regression Analyses of Lead Air Pollution Patterns and Sources in Pennsylvania"
subtitle: "https://github.com/nyb5208/Alcorn_Bao_Hermanson_ENV_872_EDA_FinalProject.git"
author: "Jack Alcorn, Max Hermanson, and Nancy Bao "
fontsize: 12pt
mainfont: Times New Roman
editor_options:
  chunk_output_type: console
---

\newpage
\tableofcontents 

\newpage
\listoftables
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory
getwd()
# Load your packages
library(tidyverse)
library(ggplot2)
library(ggridges)
library(ggpubr)
library(dplyr)
library(lubridate)
library(zoo)
library(trend)
library(sf)
library(Kendall)
library(knitr)
library(kableExtra)
#Load color palette packages
library(RColorBrewer)
library(wesanderson)
library(viridisLite)
library(viridis)
library(colormap)
library(mapview)
library(sf)
# Set your ggplot theme
project.theme <- theme_classic(base_size=12)+
                 theme(plot.title= 
                         element_text(size=14,
                         face="bold",
                         color= "black",
                         hjust=0.5),
                         axis.text = 
                         element_text(color = "black"),
                         axis.title = 
                         element_text(color= "black",face= "bold"),
                         legend.position="top")
theme_set(project.theme)
#___________________

# Load EPA Air Lead Pollution datasets
epa.PAlead.2010 <-read.csv( "./Data/Raw/EPA_Lead_PA/EPA_10_PA_lead.csv", stringsAsFactors = TRUE)
epa.PAlead.2011 <-read.csv( "./Data/Raw/EPA_Lead_PA/EPA_11_PA_lead.csv", stringsAsFactors = TRUE)
epa.PAlead.2012 <-read.csv( "./Data/Raw/EPA_Lead_PA/EPA_12_PA_lead.csv", stringsAsFactors = TRUE)
epa.PAlead.2013 <-read.csv( "./Data/Raw/EPA_Lead_PA/EPA_13_PA_lead.csv", stringsAsFactors = TRUE)
epa.PAlead.2014 <-read.csv( "./Data/Raw/EPA_Lead_PA/EPA_14_PA_lead.csv", stringsAsFactors = TRUE)
epa.PAlead.2015 <-read.csv( "./Data/Raw/EPA_Lead_PA/EPA_15_PA_lead.csv", stringsAsFactors = TRUE)
epa.PAlead.2016 <-read.csv( "./Data/Raw/EPA_Lead_PA/EPA_16_PA_lead.csv", stringsAsFactors = TRUE)
epa.PAlead.2017 <-read.csv( "./Data/Raw/EPA_Lead_PA/EPA_17_PA_lead.csv", stringsAsFactors = TRUE)
epa.PAlead.2018 <-read.csv( "./Data/Raw/EPA_Lead_PA/EPA_18_PA_lead.csv", stringsAsFactors = TRUE)
epa.PAlead.2019 <-read.csv( "./Data/Raw/EPA_Lead_PA/EPA_19_PA_lead.csv", stringsAsFactors = TRUE)
epa.PAlead.2020 <-read.csv( "./Data/Raw/EPA_Lead_PA/EPA_20_PA_lead.csv", stringsAsFactors = TRUE)
# Combined the EPA lead air data sets with rbind()
epa.PAair.lead <-rbind(epa.PAlead.2010, epa.PAlead.2011, epa.PAlead.2012, epa.PAlead.2013,
                       epa.PAlead.2014, epa.PAlead.2015, epa.PAlead.2016, epa.PAlead.2017,
                       epa.PAlead.2018, epa.PAlead.2019, epa.PAlead.2020)

#Change date variable from factor to date
epa.PAair.lead$Date <- as.Date(epa.PAair.lead$Date, format="%m/%d/%Y")
class(epa.PAair.lead$Date)
#___________________
#Read in PA Counties shp file (State FIPS = 42)
PA_county_sf <- st_read("./Data/Spatial/cb_2018_us_county_20m.shp") %>%
filter(STATEFP == 42)

```


# Rationale and Research Questions
>

- ***Question 1: How do air lead concentrations vary from 2010 to 2020 across Pennsylvania?***

- ***Question 2: Where are the hotspots for air lead concentrations in Pennsylvania?***

- ***Question 3: PA Income, resperitory disease, water lead levels, number of: airports, metal processing plants, coal plants, incinerators***


\newpage

# Dataset Information



\newpage

# Exploratory Analysis 
```{r exploratory-analysis, include= FALSE, warning = FALSE,message=FALSE}
#
dim(epa.PAair.lead)
colnames(epa.PAair.lead)
head(epa.PAair.lead)

#Cleaned and Processed EPA PA Air Lead Data
epa.PA.airPb.processed <- epa.PAair.lead %>%
                                  select(Date, Site.ID, 
                                         Daily.Mean.Pb.Concentration,
                                         Site.Name,AQS_PARAMETER_DESC,CBSA_NAME,CBSA_CODE,
                                         STATE,COUNTY, COUNTY_CODE) %>%
                                  mutate(Month = month(Date),
                                  Year=year(Date)) %>%
                                  mutate(Date_combined = my(paste0(Month,"-",Year)))
```


\newpage
# Introduction 

# Analysis
```{r time_series-data-wrangling,echo=FALSE}
#Looking at the lead air concentrations for major metropolitan areas in Pennsylvania 
# Looking at the Combined Statistical Areas (CBSA) to identify the metropolitan areas included in this study
summary(epa.PA.airPb.processed$CBSA_NAME)
summary(as.factor(epa.PA.airPb.processed$CBSA_CODE))
#associated CBSA codes 
### Allentown-Bethlehem-Easton, PA-NJ =10900
### Chambersburg-Waynesboro =16540
### Erie=21500
### Indiana=26860
### Lancaster=29540
### Lewisburg=30260
### New Castle=35260
### Reading=39740
### Pittsburgh =38300
### Philadelphia-Camden-Wilmington, PA-NJ-DE-MD =39780
### Reading=39740
summary(epa.PA.airPb.processed$COUNTY)

#Need to determine the top 5 metropolitan areas by 2010 US Census population size (most populated):
## Loaded US census data to determine the most populated metropolitan areas from the CBSAs provided in the epa Air Lead data
PA.msa.population.size <-read.csv("./Data/Raw/PA_Metro_Area_data/cbsa-est2019-alldata.csv")
#Set CBSA variable as factor
as.factor(PA.msa.population.size$CBSA)
### Process the PA.msa.population.size data so I only have PA metro areas filtered by CBSA code
PA.msa.pop.processed <- PA.msa.population.size %>%
                          filter(LSAD=="Metropolitan Statistical Area") %>%
                          select(CBSA, NAME, CENSUS2010POP) %>%
                          filter(CBSA %in% c("10900", "16540", "21500", "26860", "29540",
                                        "30260", "35260", "37980",
                                        "38300", "39740", "42540"))%>%
                          arrange(desc(CENSUS2010POP))
#Picked the Top 5 Metropolitan Areas to look at based on 2010 US Census Population Size:
## 1. Philadelphia-Camden-Wilmington, PA-NJ-DE-MD 
## 2. Pittsburgh, PA  
## 3. Allentown-Bethlehem-Easton, PA-NJ 
## 4. Scranton--Wilkes-Barre--Hazleton, PA
## 5. Lancaster, PA 
#Looked at Monthly Lead Air concentrations of the Five major metropolitan areas in Pennsylvania 
#Philadelphia-Camden-Wilmington, PA-NJ-DE-MD 
epa.PA.monthly.Pb.Philly <-epa.PA.airPb.processed %>%
                  filter(CBSA_CODE==37980)%>%
                  select(Date,Date_combined, Daily.Mean.Pb.Concentration,) %>%
                  group_by(Date_combined)%>%
                  dplyr::summarise(Mean.monthly.Pb = mean(Daily.Mean.Pb.Concentration))%>%
                  add_column(CBSA="Philadelphia-Camden-Wilmington")
epa.PA.monthly.Pb.Philly
#
#Pittsburgh Metropolitan Area
epa.PA.monthly.Pb.Pitt <-epa.PA.airPb.processed %>%
                  filter(CBSA_CODE==38300)%>%
                  select(Date,Date_combined, Daily.Mean.Pb.Concentration,) %>%
                  group_by(Date_combined)%>%
                  dplyr::summarise(Mean.monthly.Pb = mean(Daily.Mean.Pb.Concentration))%>%
                  add_column(CBSA="Pittsburgh")
epa.PA.monthly.Pb.Pitt
#only have data from 2010 to 2017 for Pittsburgh area (Allegheny Co., PA)
#
# Allentown-Bethlehem-Easton, PA-NJ Metropolitan Area
epa.PA.monthly.Pb.Allentown <-epa.PA.airPb.processed %>%
                  filter(CBSA_CODE==10900)%>%
                  select(Date,Date_combined, Daily.Mean.Pb.Concentration,) %>%
                  group_by(Date_combined)%>%
                  dplyr::summarise(Mean.monthly.Pb = mean(Daily.Mean.Pb.Concentration))%>%
                  add_column(CBSA="Allentown-Bethlehem-Easton")
epa.PA.monthly.Pb.Allentown 
#Now looking at major mining county: Scranton-Wilkes-Barre-Hazleton Metropolitan Area (Luzerne Co, PA)
epa.PA.monthly.Pb.Scranton <-epa.PA.airPb.processed %>%
                  filter(CBSA_CODE==42540)%>%
                  select(Date,Date_combined,COUNTY, Daily.Mean.Pb.Concentration,) %>%
                  group_by(Date_combined)%>%
                  dplyr::summarise(Mean.monthly.Pb = mean(Daily.Mean.Pb.Concentration))%>%
                  add_column(CBSA="Scranton-Wilkes_Barre-Hazleton")
epa.PA.monthly.Pb.Scranton 
# Lancaster Metropolitan Area 
epa.PA.monthly.Pb.Lancaster <-epa.PA.airPb.processed %>%
                  filter(CBSA_CODE==29540)%>%
                  select(Date,Date_combined,COUNTY, Daily.Mean.Pb.Concentration,) %>%
                  group_by(Date_combined)%>%
                  dplyr::summarise(Mean.monthly.Pb = mean(Daily.Mean.Pb.Concentration))%>%
                  add_column(CBSA="Lancaster")
epa.PA.monthly.Pb.Lancaster
#Created a monthly mean Pb air concentration in PA counties 
Pa.monthly.Pb.MSA<-rbind(epa.PA.monthly.Pb.Philly, epa.PA.monthly.Pb.Pitt, epa.PA.monthly.Pb.Allentown, epa.PA.monthly.Pb.Scranton, epa.PA.monthly.Pb.Lancaster)
Pa.monthly.Pb.MSA
class(Pa.monthly.Pb.MSA$Date_combined)
```


```{r time-series-visuals}
#Looking at the top 5 Metropolitan Areas for air Pb concentrations from 2010 to 2020
#PA Metropolitan Area
PAmonthly_plot<-ggplot(Pa.monthly.Pb.MSA,aes(x = Date_combined,
                      y = Mean.monthly.Pb, color= CBSA))+
                      geom_line() +
                      labs(y="Monthly Average Pb Air Concentration (ug/m^3)",
                           x="Date",
                          title="Monthly Average Pb Air Concentrations from 2010 to 2020 in PA Metro Areas")+
                          scale_x_date(date_breaks = "3 months",date_labels = "%b %Y",
                                       limits=c(as.Date("2010-01-01"), as.Date("2021-01-01")),
             expand=c(0,0))+
                          theme(axis.text.x=element_text(angle=90, hjust=1))+
                      geom_hline(yintercept=0.15, linetype='dotted', col = 'black')+
                      annotate("text", x = as.Date("2016-12-01", "%Y-%m-%d"), y = 0.15,
                               label = "EPA Primary and Secondary 3-month Average Pb NAAQ Standard=0.15ug/m^3", vjust = -0.5)
print(PAmonthly_plot)

```

```{r linear_interpolation,include=FALSE}
# Philadelphia-Camden-Wilmington, PA-NJ-DE-MD 
Philadelphia.days <-as.data.frame(seq(ymd("2010-01-02"), ymd("2020-12-29"),by="days"))
Philadelphia.days <-setNames(Philadelphia.days,c("Date"))
Phila.msa.Pb.filter <-epa.PA.airPb.processed %>%
                        filter(CBSA_CODE==37980) %>%
                        select(Date,Daily.Mean.Pb.Concentration)
Phila.msa.Pb<-left_join(Philadelphia.days,Phila.msa.Pb.filter)
Phila.msa.Pb.cleaned <-Phila.msa.Pb %>%
                        mutate(Daily.Mean.Pb.Concentration.clean=
                                 zoo::na.approx(Daily.Mean.Pb.Concentration)) %>%
                        add_column(MSA=c("Philadelphia-Camden-Wilmington, PA-NJ-DE-MD"))                        
# Pittsburgh, PA 
Pitt.days <-as.data.frame(seq(ymd("2010-01-02"), ymd("2020-12-29"),by="days"))
Pitt.days <-setNames(Pitt.days, c("Date"))
Pitt.msa.Pb.filter<-epa.PA.airPb.processed %>%
                        filter(CBSA_CODE==38300) %>%
                        select(Date,Daily.Mean.Pb.Concentration)
Pitt.msa.Pb<-left_join(Pitt.days,Pitt.msa.Pb.filter)
Pitt.msa.Pb.cleaned <- Pitt.msa.Pb %>%
                        mutate(Daily.Mean.Pb.Concentration.clean=
                                 zoo::na.approx(Daily.Mean.Pb.Concentration)) %>%
                        add_column(MSA=c("Pittsburgh, PA"))
# Allentown-Bethlehem-Easton, PA-NJ 
Allentown.days <-as.data.frame(seq(ymd("2012-07-02"), ymd("2020-12-29"),by="days"))
Allentown.days<-setNames(Allentown.days,c("Date"))
Allentown.msa.Pb.filter<-epa.PA.airPb.processed %>%
                   filter(CBSA_CODE==10900)%>%
                   select(Date,Daily.Mean.Pb.Concentration)
Allentown.msa.Pb<-left_join(Allentown.days,Allentown.msa.Pb.filter)
#Cleaned Allentown data
Allentown.msa.Pb.cleaned <-Allentown.msa.Pb %>%
  mutate(Daily.Mean.Pb.Concentration.clean=
           zoo::na.approx(Daily.Mean.Pb.Concentration))%>%
           add_column(MSA=c("Allentown-Bethlehem-Easton, PA-NJ"))
Allentown.msa.Pb.cleaned
#
# Scranton--Wilkes-Barre--Hazleton, PA
Scranton.days <-as.data.frame(seq(ymd("2010-01-02"), ymd("2018-09-29"),by="days"))
Scranton.days <-setNames(Scranton.days,c("Date"))
Scranton.msa.Pb.filter<-epa.PA.airPb.processed %>%
                          filter(CBSA_CODE==42540)%>%
                          select(Date,Daily.Mean.Pb.Concentration)
Scranton.msa.Pb<-left_join(Scranton.days,Scranton.msa.Pb.filter)
#Cleaned Scranton data
Scranton.msa.Pb.cleaned <-Scranton.msa.Pb %>%
  mutate(Daily.Mean.Pb.Concentration.clean=
           zoo::na.approx(Daily.Mean.Pb.Concentration)) %>%
           add_column(MSA=c("Scranton--Wilkes-Barre--Hazleton, PA"))
Scranton.msa.Pb.cleaned
#
#Lancaster, PA
Lancaster.days <-as.data.frame(seq(ymd("2012-01-04"), ymd("2020-12-29"),by="days"))
Lancaster.msa.Pb.filter<- epa.PA.airPb.processed %>%
                          filter(CBSA_CODE=="29540") %>%
                          select(Date,Daily.Mean.Pb.Concentration)

Lancaster.days<-setNames(Lancaster.days,c("Date"))
#
Lancaster.msa.Pb <-left_join(Lancaster.days,Lancaster.msa.Pb.filter)
Lancaster.msa.Pb.cleaned <-Lancaster.msa.Pb %>%
  mutate(Daily.Mean.Pb.Concentration.clean=
           zoo::na.approx(Daily.Mean.Pb.Concentration))%>%
           add_column(MSA=c("Lancaster, PA"))
```

```{r linear-interp-visual,echo=FALSE}
#first I joined all the linear interpolated data together
PA.msa.Pb.cleaned <-rbind(Phila.msa.Pb.cleaned,Pitt.msa.Pb.cleaned,Allentown.msa.Pb.cleaned,Scranton.msa.Pb.cleaned, Lancaster.msa.Pb.cleaned)
class(PA.msa.Pb.cleaned$Date)
#
PA.msa.Pb.cleaned.plot<-ggplot(PA.msa.Pb.cleaned, aes(x = Date,
                      y = Daily.Mean.Pb.Concentration.clean, color= MSA))+
                      geom_line() +
                      labs(y="Daily Average Air Pb Level (ug/m^3)",
                           x="Date",
                          title="Daily Average Air Pb Levels from 2010 to 2020 in Metropolitan Areas in PA",
                          color="")+
                          scale_x_date(date_breaks = "3 months",date_labels = "%b %Y",
                                       limits=c(as.Date("2010-01-01"), as.Date("2021-01-01")),
             expand=c(0,0))+
                          theme(axis.text.x=element_text(angle=90, hjust=1))+
                      geom_hline(yintercept=0.15, linetype='dotted', col = 'black')+
                      annotate("text", x = as.Date("2016-12-01", "%Y-%m-%d"), y = 0.15,
                               label = "EPA Primary and Secondary 3-month Average Pb NAAQ Standard=0.15ug/m^3", vjust = -0.5)
print(PA.msa.Pb.cleaned.plot)

```

```{r time-series-analysis}
#I am looking at the  metropolitan areas with Air Pb Levels that exceeded the EPA primary and secondary NAAQs
#Pittsburgh
Pitt.msa.linear.interp.plot <-ggplot(Pitt.msa.Pb.cleaned)+
geom_line(aes(x = Date,
y = Daily.Mean.Pb.Concentration.clean),
color = "chartreuse4", alpha=0.5) +
labs(y="Daily Mean Air Pb Level (ug/m^3)",title="Daily Mean Air Pb Levels in the Pittsburgh, PA Metropolitan Area from 2010 to 2020")+
  scale_x_date(date_breaks = "3 months",date_labels = "%b %Y",
                                       limits=c(as.Date("2010-01-01"), as.Date("2020-01-01")),
             expand=c(0,0))+
                      geom_hline(yintercept=0.15, linetype='dotted', col = 'black')+
                      annotate("text", x = as.Date("2016-12-01", "%Y-%m-%d"), y = 0.15,
                               label = "EPA Primary and Secondary 3-month Average Pb NAAQ Standard=0.15ug/m^3", vjust = -0.5)+
                          theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(plot.margin=unit(c(1,1,1.5,1.2),"cm"))
print(Pitt.msa.linear.interp.plot)
#
Pitt.monthly.Pb.msa <-Pitt.msa.Pb.cleaned %>%
                        mutate(Month = month(Date),
                        Year=year(Date)) %>%
                        mutate(Date_combined = my(paste0(Month,"-",Year)))%>%
                        group_by(Date_combined)%>%
                        dplyr::summarise(Mean.monthly.Pb=mean(Daily.Mean.Pb.Concentration.clean))
Pitt.monthly.Pb.msa <- Pitt.monthly.Pb.msa %>%
                           mutate(Date=as.Date(Date_combined))
#Checking the first and last dates of data collection
first(Pitt.msa.Pb.cleaned$Date)
last(Pitt.msa.Pb.cleaned$Date)
#
#Time Series: daily
Pitt.msa.Pb.daily.ts <-ts(Pitt.msa.Pb.cleaned$Daily.Mean.Pb.Concentration.clean,
                              start=c(2010,1),
                              end=c(2020,12),
                              frequency=365)
#Time Series: monthly
Pitt.msa.Pb.monthly.ts <-ts(Pitt.monthly.Pb.msa$Mean.monthly.Pb,
                                start=c(2010,1),
                                end=c(2020,12),
                                frequency=12)
#
Pitt.msa.Pb.daily.decomp <- stl(Pitt.msa.Pb.daily.ts, s.window = "periodic")
plot(Pitt.msa.Pb.daily.decomp)

Pitt.msa.Pb.monthly.decomp <-stl(Pitt.msa.Pb.monthly.ts, s.window = "periodic")
plot(Pitt.msa.Pb.monthly.decomp)
#
Pitt.Pb.daily.trend.nonseasonal <-Kendall::MannKendall(Pitt.msa.Pb.daily.ts)
summary(Pitt.Pb.daily.trend.nonseasonal)
#
Pitt.Pb.monthly.trend.nonseasonal <-Kendall::MannKendall(Pitt.msa.Pb.monthly.ts)
summary(Pitt.Pb.monthly.trend.nonseasonal)
#Scranton-Wilkes-Barre-Hazleton
#Scranton linear interpolation
Scranton.msa.linear.interpolation<-ggplot(Scranton.msa.Pb.cleaned) +
geom_line(aes(x = Date,
y = Daily.Mean.Pb.Concentration.clean),
color = "blue", alpha=0.5) +
labs(y="Daily Mean Pb Air Concentration (ug/m^3)",title="Daily Mean Air Pb Levels in the Scranton-Wilkes-Barre-Hazleton Metropolitan Area from 2010 to 2018")+
  scale_x_date(date_breaks = "3 months",date_labels = "%b %Y",
                                       limits=c(as.Date("2010-01-01"), as.Date("2018-10-01")),
             expand=c(0,0))+
                      geom_hline(yintercept=0.15, linetype='dotted', col = 'black')+
                      annotate("text", x = as.Date("2014-10-01", "%Y-%m-%d"), y = 0.15,
                               label = "EPA Primary and Secondary 3-month Average Pb NAAQ Standard=0.15ug/m^3", vjust = -0.5)+
                          theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(plot.margin=unit(c(1,1,1.5,1.2),"cm"))
print(Scranton.msa.linear.interpolation)
#
Scranton.monthly.Pb.msa <-Scranton.msa.Pb.cleaned %>%
                        mutate(Month = month(Date),
                        Year=year(Date)) %>%
                        mutate(Date_combined = my(paste0(Month,"-",Year)))%>%
                        group_by(Date_combined)%>%
                        dplyr::summarise(Mean.monthly.Pb=mean(Daily.Mean.Pb.Concentration.clean))
Scranton.monthly.Pb.msa <- Scranton.monthly.Pb.msa %>%
                           mutate(Date=as.Date(Date_combined))
#Checking the first and last dates of data collection
first(Scranton.msa.Pb.cleaned$Date)
last(Scranton.msa.Pb.cleaned$Date)
#
#Time Series: daily
Scranton.msa.Pb.daily.ts <-ts(Scranton.msa.Pb.cleaned$Daily.Mean.Pb.Concentration.clean,
                              start=c(2010,1),
                              end=c(2018,9),
                              frequency=365)
#Time Series: monthly
Scranton.msa.Pb.monthly.ts <-ts(Scranton.monthly.Pb.msa$Mean.monthly.Pb,
                                start=c(2010,1),
                                end=c(2018,9),
                                frequency=12)
#
Scranton.msa.Pb.daily.decomp <- stl(Scranton.msa.Pb.daily.ts, s.window = "periodic")
plot(Scranton.msa.Pb.daily.decomp)

Scranton.msa.Pb.monthly.decomp <-stl(Scranton.msa.Pb.monthly.ts, s.window = "periodic")
plot(Scranton.msa.Pb.monthly.decomp)
#
Scranton.Pb.daily.trend.nonseasonal <-Kendall::MannKendall(Scranton.msa.Pb.daily.ts)
summary(Scranton.Pb.daily.trend.nonseasonal)
#
Scranton.Pb.monthly.trend.nonseasonal <-Kendall::MannKendall(Scranton.msa.Pb.monthly.ts)
summary(Scranton.Pb.monthly.trend.nonseasonal)
#Lancaster, PA
#Lancaster linear interpolation
Lancaster.msa.linear.interpolation<-ggplot(Lancaster.msa.Pb.cleaned) +
geom_line(aes(x = Date,
y = Daily.Mean.Pb.Concentration.clean),
color = "orange", alpha=1)
labs(y="Daily Mean Air Pb Level (ug/m^3)",title="Daily Mean Air Pb Levels in the Lancaster, PA Metropolitan Area from 2012 to 2020")+
  scale_x_date(date_breaks = "3 months",date_labels = "%b %Y",
                                       limits=c(as.Date("2012-01-01"), as.Date("2021-01-01")),
             expand=c(0,0))+
                      geom_hline(yintercept=0.15, linetype='dotted', col = 'black')+
                      annotate("text", x = as.Date("2014-10-01", "%Y-%m-%d"), y = 0.15,
                               label = "EPA Primary and Secondary 3-month Average Pb NAAQ Standard=0.15ug/m^3", vjust = -0.5)+
                          theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(plot.margin=unit(c(1,1,1.5,1.2),"cm"))
print(Lancaster.msa.linear.interpolation)
#
Lancaster.monthly.Pb.msa <-Lancaster.msa.Pb.cleaned %>%
                        mutate(Month = month(Date),
                        Year=year(Date)) %>%
                        mutate(Date_combined = my(paste0(Month,"-",Year)))%>%
                        group_by(Date_combined)%>%
                        dplyr::summarise(Mean.monthly.Pb=mean(Daily.Mean.Pb.Concentration.clean))
Lancaster.monthly.Pb.msa <- Lancaster.monthly.Pb.msa %>%
                           mutate(Date=as.Date(Date_combined))
#Checking the first and last dates of data collection
first(Lancaster.msa.Pb.cleaned$Date)
last(Lancaster.msa.Pb.cleaned$Date)
#
#Time Series: daily
Lancaster.msa.Pb.daily.ts <-ts(Lancaster.msa.Pb.cleaned$Daily.Mean.Pb.Concentration.clean,
                              start=c(2012,1),
                              end=c(2020,12),
                              frequency=365)
#Time Series: monthly
Lancaster.msa.Pb.monthly.ts <-ts(Lancaster.monthly.Pb.msa$Mean.monthly.Pb,
                                start=c(2012,1),
                                end=c(2020,12),
                                frequency=12)
#
Lancaster.msa.Pb.daily.decomp <- stl(Lancaster.msa.Pb.daily.ts, s.window = "periodic")
plot(Lancaster.msa.Pb.daily.decomp)

Lancaster.msa.Pb.monthly.decomp <-stl(Lancaster.msa.Pb.monthly.ts, s.window = "periodic")
plot(Lancaster.msa.Pb.monthly.decomp)
#
Lancaster.Pb.daily.trend.nonseasonal <-Kendall::MannKendall(Lancaster.msa.Pb.daily.ts)
summary(Lancaster.Pb.daily.trend.nonseasonal)
#
Lancaster.Pb.monthly.trend.nonseasonal <-Kendall::MannKendall(Lancaster.msa.Pb.monthly.ts)
summary(Lancaster.Pb.monthly.trend.nonseasonal)
Lancaster.Pb.monthly.trend.seasonal<-
  Kendall::SeasonalMannKendall(Lancaster.msa.Pb.monthly.ts)
summary(Lancaster.Pb.monthly.trend.seasonal)
#Allentown
Allentown.msa.linear.interpolation<-ggplot(Allentown.msa.Pb.cleaned) +
geom_line(aes(x = Date,
y = Daily.Mean.Pb.Concentration.clean),
color = "aquamarine", alpha=1) +
labs(y="Daily Mean Air Pb Level (ug/m^3)",title="Daily Mean Air Pb Levels in the Allentown-Bethlehem-Easton, PA-NJ Metropolitan Area from 2012 to 2020")+
  scale_x_date(date_breaks = "3 months",date_labels = "%b %Y",
                                       limits=c(as.Date("2012-07-01"), as.Date("2021-01-01")),
             expand=c(0,0))+
                      geom_hline(yintercept=0.15, linetype='dotted', col = 'black')+
                      annotate("text", x = as.Date("2014-10-01", "%Y-%m-%d"), y = 0.15,
                               label = "EPA Primary and Secondary 3-month Average Pb NAAQ Standard=0.15ug/m^3", vjust = -0.5)+
                          theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(plot.margin=unit(c(1,1,1.5,1.2),"cm"))
print(Allentown.msa.linear.interpolation)
#Checking the first and last dates of data collection
first(Allentown.msa.Pb.cleaned$Date)
last(Allentown.msa.Pb.cleaned$Date)
#
Allentown.monthly.Pb.msa <-Allentown.msa.Pb.cleaned %>%
                        mutate(Month = month(Date),
                        Year=year(Date)) %>%
                        mutate(Date_combined = my(paste0(Month,"-",Year)))%>%
                        group_by(Date_combined)%>%
                        dplyr::summarise(Mean.monthly.Pb=mean(Daily.Mean.Pb.Concentration.clean))
Allentown.monthly.Pb.msa <- Allentown.monthly.Pb.msa %>%
                           mutate(Date=as.Date(Date_combined))
#Time Series: daily
Allentown.msa.Pb.daily.ts <-ts(Allentown.msa.Pb.cleaned$Daily.Mean.Pb.Concentration.clean,
                              start=c(2012,7),
                              end=c(2020,12),
                              frequency=365)
Allentown.Pb.daily.trend.nonseasonal <-Kendall::MannKendall(Allentown.msa.Pb.daily.ts)
summary(Allentown.Pb.daily.trend.nonseasonal)
#Time Series: monthly
Allentown.msa.Pb.monthly.ts <-ts(Allentown.monthly.Pb.msa$Mean.monthly.Pb,
                                start=c(2012,7),
                                end=c(2020,12),
                                frequency=12)
Allentown.Pb.monthly.trend.nonseasonal <- Kendall::MannKendall(Allentown.msa.Pb.monthly.ts)
summary(Allentown.Pb.monthly.trend.nonseasonal)
#Decomposition of the time series
Allentown.msa.Pb.daily.decomp <- stl(Allentown.msa.Pb.daily.ts, s.window = "periodic")
plot(Allentown.msa.Pb.daily.decomp)
#
Allentown.msa.Pb.monthly.decomp <- stl(Allentown.msa.Pb.monthly.ts, s.window = "periodic")
plot(Allentown.msa.Pb.monthly.decomp)
#Decided to look at Philadelphia as well
#Checking the first and last dates of data collection
first(Phila.msa.Pb.cleaned$Date)
last(Phila.msa.Pb.cleaned$Date)
#
Phila.monthly.Pb.msa <-Phila.msa.Pb.cleaned %>%
                        mutate(Month = month(Date),
                        Year=year(Date)) %>%
                        mutate(Date_combined = my(paste0(Month,"-",Year)))%>%
                        group_by(Date_combined)%>%
                        dplyr::summarise(Mean.monthly.Pb=mean(Daily.Mean.Pb.Concentration.clean))
Phila.monthly.Pb.msa <- Phila.monthly.Pb.msa %>%
                           mutate(Date=as.Date(Date_combined))
#Time Series: daily
Phila.msa.Pb.daily.ts <-ts(Phila.msa.Pb.cleaned$Daily.Mean.Pb.Concentration.clean,
                              start=c(2010,1),
                              end=c(2020,12),
                              frequency=365)
Phila.Pb.daily.trend.nonseasonal <-Kendall::MannKendall(Phila.msa.Pb.daily.ts)
summary(Phila.Pb.daily.trend.nonseasonal)
#Time Series: monthly
Phila.msa.Pb.monthly.ts <-ts(Phila.monthly.Pb.msa$Mean.monthly.Pb,
                                start=c(2010,1),
                                end=c(2020,12),
                                frequency=12)
Phila.Pb.monthly.trend.nonseasonal <- Kendall::MannKendall(Phila.msa.Pb.monthly.ts)
summary(Phila.Pb.monthly.trend.nonseasonal)
#Decomposition of the time series
Phila.msa.Pb.daily.decomp <- stl(Phila.msa.Pb.daily.ts, s.window = "periodic")
plot(Phila.msa.Pb.daily.decomp)
#
Phila.msa.Pb.monthly.decomp <- stl(Phila.msa.Pb.monthly.ts, s.window = "periodic")
plot(Phila.msa.Pb.monthly.decomp)
```

```{r spatial-analysis}
#Reduce our data to just one record for each location, computing mean and max daily AQI values 2010 and 2020

EPAair_PA_Lead_avg2010 <- epa.PAlead.2010 %>%
  group_by(COUNTY, SITE_LATITUDE, SITE_LONGITUDE) %>%
  summarize(meanPb = mean(Daily.Mean.Pb.Concentration),
            maxPb = max(Daily.Mean.Pb.Concentration))


EPAair_PA_Lead_avg2020 <- epa.PAlead.2020 %>%
  group_by(COUNTY, SITE_LATITUDE, SITE_LONGITUDE) %>%
  summarize(meanPb = mean(Daily.Mean.Pb.Concentration),
            maxPb = max(Daily.Mean.Pb.Concentration))

#Convert the dataset to a spatially enabled "sf" data frame
sf_PA_Lead_avg2010 <- st_as_sf(EPAair_PA_Lead_avg2010,
                             coords = c('SITE_LONGITUDE','SITE_LATITUDE'),
                             crs=4326)

sf_PA_Lead_high_2010<-filter(sf_PA_Lead_avg2010, maxPb > 0.15)
range(sf_PA_Lead_high_2010$meanPb)



sf_PA_Lead_avg2020 <- st_as_sf(EPAair_PA_Lead_avg2020,
                             coords = c('SITE_LONGITUDE','SITE_LATITUDE'),
                             crs=4326)

sf_PA_Lead_high_2020<-filter(sf_PA_Lead_avg2020, maxPb > 0.15)

#read in county shape file
counties_sf<- st_read('./Data/Spatial/cb_2018_us_county_20m.shp')%>%
  filter(STATEFP == 42) #Filter for just PA Counties
colnames(counties_sf)[2]<-"CNTY_FIPS"
colnames(counties_sf)[6]<-"COUNTY"

counties_sf$CNTY_FIPS<-as.numeric(counties_sf$CNTY_FIPS)


#read in CDC Data for each year and group by county
#2010
CDC_PA_2010<-read.csv("./Data/Raw/Pennsylvania_CDC_2010.csv", header = TRUE)

CDC_PA_2010$CNTY_FIPS<-as.numeric(CDC_PA_2010$CNTY_FIPS)

CDC_PA_2010_avg<-CDC_PA_2010%>%
  group_by(CNTY_FIPS) %>%
  summarize(meanPCI = mean(E_PCI),
            meanPOV = mean(E_P_POV),
            meanPOP = sum(TOTPOP))

range(CDC_PA_2010_avg$meanPCI)

#join CDC data and county shape file
PA_county_2010<-left_join(counties_sf, CDC_PA_2010_avg, by = c("CNTY_FIPS"))


ggplot() + 
  geom_sf(data=PA_county_2010,aes(fill = meanPOV),alpha=0.3) + 
  scale_fill_gradient2(low="blue",high="red", midpoint = .15) +
   labs(fill = "% Pop in Poverty")+
  geom_sf(data=sf_PA_Lead_avg2010, aes(size=meanPb))+
   labs(size="Lead Values
       (µg/m3)")+
  ggtitle("2010 Mean Lead Values Across Pennsylvania")

ggplot() + 
  geom_sf(data=PA_county_2010,aes(fill = meanPCI),alpha=0.3) + 
  scale_fill_gradient2(low="red",high="blue", midpoint = 28000, breaks=c(16000,28000, 40000)) + 
  labs(fill = "Per Capita 
    Income")+
  geom_sf(data=sf_PA_Lead_avg2010, aes(size=meanPb))+
   labs(size="Lead Values
       (µg/m3)")+
  ggtitle("2010 Mean Lead Values Across Pennsylvania")

ggplot() + 
  geom_sf(data=PA_county_2010,aes(fill = meanPCI),alpha=0.3) + 
  scale_fill_gradient(low="blue", high="red", breaks=c(16000,28000, 40000)) +
  labs(fill="Per Capita Income")+
  geom_sf(data=sf_PA_Lead_high_2010, aes(size=maxPb))+
   labs(size="Lead Values
       (µg/m3)")+
  ggtitle("2010 Max Lead Levels over .15 (µg/m3)")






#2018
CDC_PA_2018<-read.csv("./Data/Raw/Pennsylvania_CDC_2018.csv", header = TRUE)

CDC_PA_2018$CNTY_FIPS<-as.numeric(CDC_PA_2018$CNTY_FIPS)

CDC_PA_2018_avg<-CDC_PA_2018%>%
  group_by(COUNTY) %>%
  summarize(meanPCI = mean(E_PCI),
            meanPOV = mean(EP_POV),
            meanPOP = sum(E_TOTPOP))

CDC_PA_2018_avg<-mutate(CDC_PA_2018_avg, meanPOV = (meanPOV/100))
CDC_PA_2018_avg$CNTY_FIPS<-CDC_PA_2010_avg$CNTY_FIPS
#join CDC data and county shape file
PA_county_2018<-left_join(counties_sf, CDC_PA_2018_avg, by = c("CNTY_FIPS"))

ggplot() + 
  geom_sf(data=PA_county_2018,aes(fill = meanPOV),alpha=0.3) + 
   scale_fill_gradient2(low="blue",high="red", midpoint = .15) +
   labs(fill = "% Pop in Poverty")+
  geom_sf(data=sf_PA_Lead_avg2020, aes(size=meanPb))+
  labs(size="Lead Values 
       (µg/m3)")+
  ggtitle("2020 Mean Lead Values Across Pennsylvania")

ggplot() + 
  geom_sf(data=PA_county_2018,aes(fill = meanPCI),alpha=0.3) + 
   scale_fill_gradient2(low="red",high="blue", midpoint = 35000, breaks=c(21000,35000, 47000)) +
  labs(fill = "Per Capita
  Income")+
  geom_sf(data=sf_PA_Lead_avg2020, aes(size=meanPb))+
  labs(size="Lead Values
       (µg/m3)")+
  ggtitle("2020 Mean Lead Values Across Pennsylvania")

ggplot() + 
  geom_sf(data=PA_county_2018,aes(fill = meanPCI),alpha=0.3) + 
  scale_fill_gradient2(low="red",high="blue", midpoint = 35000, breaks=c(20000,35000, 45000)) +
  labs(fill="Per Capita Income")+
  geom_sf(data=sf_PA_Lead_high_2020, aes(size=maxPb))+
   labs(size="Lead Values
       (µg/m3)")+
  ggtitle("2020 Max Lead Levels over .15 (µg/m3)")



```

```{r Regression Data Wrangling}
#
epa.PAcounty.Pb2 <- epa.PA.airPb.processed %>%    #ADDED BY MAX FOR REGRESSION STUFF 
   select(Date,COUNTY, Daily.Mean.Pb.Concentration, COUNTY_CODE) %>%
                  group_by(COUNTY_CODE) %>%
  summarise(County.Mean.Pb=mean(Daily.Mean.Pb.Concentration))%>%
                  arrange(desc(County.Mean.Pb))

###2010 Census #####
census_sf <- st_read('./Data/Raw/CensusData/tl_2020_42_county10/tl_2020_42_county10.shp')
mapview(census_sf)
census_sf$FIPS <- census_sf$STATEFP10

###Metal####
#need to use R spatial analysis to determine which county FIPS each metal processing plant is in 
metal_sf <- st_read('./Data/Raw/Metal_processing/PA_DEP/Industrial_Mineral_Mining_Operations_(IMMO)_-_Mineral_Preparation_Plant/Industrial_Mineral_Mining_Operations_(IMMO)_-_Mineral_Preparation_Plant.shp')
metal_csv <- read.csv("./Data/Raw/Metal_processing/PA_DEP/Industrial_Mineral_Mining_Operations_(IMMO)_-_Mineral_Preparation_Plant.csv")

census_sf <- st_read('./Data/Raw/CensusData/tl_2020_42_county10/tl_2020_42_county10.shp')
mapview(census_sf)
census_sf$FIPS <- census_sf$STATEFP10
st_crs(census_sf)$epsg
metal_data_sf <- st_as_sf(metal_csv, coords = c('X', 'Y'), crs = 4269)
metal_4269 <- st_transform(metal_sf, crs = 4269)

class(metal_4269)
metal_join <- st_join(metal_4269, census_sf, join = st_intersects)
metal_join
class(metal_join$COUNTYFP10)
metal_join$COUNTYFP10

metal_summary <- metal_join %>% 
  filter(SITE_STATU == "ACTIVE") %>% 
  select(COUNTYFP10) %>% 
  count(COUNTYFP10) %>% 
  mutate(number_metal_plants = n) %>% 
  mutate(COUNTYFP10 = sub("^0+", "", COUNTYFP10)) %>% #get rid of trailing zeros 
  select(COUNTYFP10, number_metal_plants, geometry)
census_sf2 <- census_sf %>% 
  select(COUNTYFP10, NAME10)

metal_join_census <- st_join(census_sf2, metal_summary)  ## geographic joining 
metal_join_census[is.na(metal_join_census)] = 0  #remove NAs 
metal_join_census <- metal_join_census %>% 
  select(COUNTYFP10.x, NAME10, n, geometry) 

metal_join_census$dummy_metal <- as.numeric(metal_join_census$n > 0)



#####airport#####
airport_raw <- read.csv("./Data/Raw/Airport Data/Airports.csv",  stringsAsFactors = TRUE)
airport_sf <- st_as_sf(airport_raw, coords = c('X', 'Y'), crs = 4269)
airport_PA <- airport_sf %>% 
  filter(STATE == "PA") %>% 
  filter(TYPE_CODE == "AD") %>% 
  filter(PRIVATEUSE == 0)
mapview(airport_PA)

airport_PA2_subset <- airport_PA %>% 
  select(NAME, geometry)

airport_join <- st_join(airport_PA2_subset, census_sf, join = st_intersects)
class(airport_join$COUNTYFP10)
airport_summary <- airport_join %>% 
  select(COUNTYFP10) %>% 
  count(COUNTYFP10) %>% 
  mutate(number_airports = n) %>% 
  mutate(COUNTYFP10 = sub("^0+", "", COUNTYFP10)) %>% 
  select(COUNTYFP10, number_airports, geometry)



census_sf2 <- census_sf %>% 
  select(COUNTYFP10, NAME10)
airport_join_census <- st_join(census_sf2, airport_summary)
airport_join_census$number_airports <- airport_join_census$n



metal_join_census_airport <- st_join(metal_join_census, airport_PA)




joint_nonSF <- metal_join_census %>% 
  select(COUNTYFP10.x, NAME10, n)
joint_nonSF <- joint_nonSF %>% st_drop_geometry()
joint_nonSF$COUNTYFP10 <- joint_nonSF$COUNTYFP10.x
class(joint_nonSF)

#####Incinerators #####
incinerate_sf_raw <- st_read('./Data/Raw/Incinerator_Shp_Data/Air_Emission_Plant_-_Incinerator.shp')
incinerate_sf <- st_transform(incinerate_sf_raw, crs = 4269)
st_crs(incinerate_sf)$epsg 
st_crs(census_sf2)$epsg 
incinerate_sf2 <- incinerate_sf %>% 
  select( SITE_NAME, SITE_STATU, geometry, SUB_FACILI) %>% 
  filter(SITE_STATU == "ACTIVE") %>% 
  filter(SUB_FACILI == "INCINERATOR" ) %>% 
incinerate_join_census<- st_join(incinerate_sf, census_sf2, join = st_intersects) # join 
incinerate_join_census[is.na(incinerate_join_census)] = 0  #remove NAs 
incinerate_join_census <- metal_join_census %>%  #clean data 
  select(COUNTYFP10.x, NAME10, n, geometry) 

incinerator_summary <- incinerate_join_census %>% 
  select(COUNTYFP10) %>% 
  count(COUNTYFP10) %>% 
  mutate(number_incinerators = n) %>% #rename count column
  mutate(COUNTYFP10 = sub("^0+", "", COUNTYFP10)) %>% #get rid of trailing zeros 
  select(COUNTYFP10, number_incinerators, geometry) #clean

incinerator_summary[is.na(incinerator_summary)] = 0  #remove NAs 


###Blood_Lead_Children####
##contains county FIPS code
children_blood_lead <- read.csv("./Data/Raw/BloodLeadLevels/PA_CountyLevelSummary_2017.csv", stringsAsFactors = TRUE)
children_blood_lead$COUNTYFP10<- as.character(children_blood_lead$County.FIPS)
class(children_blood_lead$COUNTYFP10)
print(children_blood_lead$COUNTYFP10)

children_blood_lead$Percent_.5 <- as.numeric(children_blood_lead$Percent_.5)
blood_lead <- children_blood_lead %>% 
  select(COUNTYFP10, County.Name, Percent_.5)


joint_nonSF$COUNTYFP10 <- sub("^0+", "", joint_nonSF$COUNTYFP10) #get rid of trailing 0's 
print(joint_nonSF$COUNTYFP10)
joint_nonSF2 <- left_join(joint_nonSF, blood_lead)
#### Joining Data ####
blood_lead #non-sf; join below three and then drop geometry, then join all. 
incinerator_summary #sf
airport_summary #sf
metal_summary #sf 
CDC_PA_2010_avg2 <- CDC_PA_2010_avg %>% 
  select(CNTY_FIPS, meanPCI, meanPOV) %>% 
  mutate(COUNTYFP10 = CNTY_FIPS) %>% 
  select(COUNTYFP10, meanPCI, meanPOV)
CDC_PA_2010_avg2$COUNTYFP10 <- as.character(CDC_PA_2010_avg2$COUNTYFP10)

class(CDC_PA_2010_avg2$COUNTYFP10)
class(i_a_m_b_join$COUNTYFP10)
incinerator_summary2 <- incinerator_summary %>% st_drop_geometry()
airport_summary2 <- airport_summary %>% st_drop_geometry()
metal_summary2 <- metal_summary %>% st_drop_geometry()  


i_a_join<- left_join(incinerator_summary2, airport_summary2)
i_a_m_join <- left_join(i_a_join, metal_summary2)
i_a_m_b_join <- left_join(i_a_m_join, blood_lead)   # join of all datasets USE IN REGRESSION 
i_a_m_b_join <- left_join(i_a_m_b_join, CDC_PA_2010_avg2)
i_a_m_b_join[is.na(i_a_m_b_join)] = 0  #remove NAs 

### Data Visualization ####
p1<-ggplot(i_a_m_b_join, aes(x=number_incinerators))+
  geom_histogram(fill="lightgreen", color="darkgreen", binwidth=1)+
  labs(x="# of Incinerators", title="Incinerators per county, histogram")
p2<-ggplot(i_a_m_b_join, aes(x=number_airports))+
  geom_histogram(fill="lightgreen", color="darkgreen", binwidth=1)+
  labs(x="# of airports", title="Airports per county, histogram")
p3<-ggplot(i_a_m_b_join, aes(x=number_metal_plants))+
  geom_histogram(fill="lightgreen", color="darkgreen", binwidth=1)+
  labs(x="# of metal plants", title="Processing plants per county, histogram")
p1
p2
p3

#### Regression Modeling ##### 
i_a_m_b_join$number_incinerators <-i_a_m_b_join$number_incinerators+0.001 #account for zeros in log
i_a_m_b_join$number_airports <- i_a_m_b_join$number_airports+0.001 #account for zeros in log
i_a_m_b_join$number_metal_plants <- i_a_m_b_join$number_metal_plants+1 #account for zeros in log

plot(x = i_a_m_b_join$number_metal_plants, y = i_a_m_b_join$Percent_.5) # consider logging
plot(x = i_a_m_b_join$number_incinerators, y = i_a_m_b_join$Percent_.5) # consider logging 
plot(x = i_a_m_b_join$number_airports, y = i_a_m_b_join$Percent_.5)
plot(x = i_a_m_b_join$meanPCI, y = i_a_m_b_join$Percent_.5)
plot(x = i_a_m_b_join$meanPOV, y = i_a_m_b_join$Percent_.5)

i_a_m_b_join$log_metal <- log(i_a_m_b_join$number_metal_plants) #log transform X
i_a_m_b_join$log_incinerate <- log(i_a_m_b_join$number_incinerators) #log transform X
i_a_m_b_join$log_blood <- log(i_a_m_b_join$Percent_.5) #log transform Y 
i_a_m_b_join$log_PCI <- log(i_a_m_b_join$meanPCI)#log transform X
i_a_m_b_join$quadPOV <- (i_a_m_b_join$meanPOV)^2# quadratic X 

plot(x = i_a_m_b_join$log_metal, y = i_a_m_b_join$Percent_.5) #log metal; looks better 
plot(x = i_a_m_b_join$log_incinerate, y = i_a_m_b_join$Percent_.5) #log incinerator ; looks better
plot(x = i_a_m_b_join$number_airports, y = i_a_m_b_join$log_blood) 
plot(x = i_a_m_b_join$log_PCI, y = i_a_m_b_join$Percent_.5)
plot(x = i_a_m_b_join$quadPOV, y = i_a_m_b_join$Percent_.5)



model1 <- lm(formula = Percent_.5 ~ number_metal_plants + number_incinerators + number_airports, data = i_a_m_b_join)
summary(model1)

model2 <- lm(formula = Percent_.5 ~ log_metal + log_incinerate + number_airports, data = i_a_m_b_join)
summary(model2)

model3 <- lm(formula = log_blood ~ number_metal_plants + number_incinerators + number_airports, data = i_a_m_b_join)
summary(model3)
#additional variables
model4 <- lm(formula = Percent_.5 ~ number_metal_plants + number_incinerators + number_airports + meanPCI + meanPOV, data = i_a_m_b_join)
summary(model4)

model5 <- lm(formula = Percent_.5 ~ number_metal_plants + number_incinerators + number_airports + log_PCI + quadPOV, data = i_a_m_b_join)
summary(model5)
plot(model5)

model6 <- lm(formula = Percent_.5 ~ number_metal_plants + log_metal+ log_incinerate + meanPCI +    number_incinerators + number_airports + log_PCI + quadPOV, data = i_a_m_b_join) # plugged in every variable and then ran through "step" function , which produced model 7 
summary(model6)
step(model6) 
#step function 

model7 <- lm(formula = Percent_.5 ~ meanPOV+ quadPOV, data = i_a_m_b_join)
summary(model7)

model8 <- lm(formula = Percent_.5 ~ log_metal + log_incinerate + number_airports + meanPCI + quadPOV, data = i_a_m_b_join)
summary(model8)

model9 <- lm(formula = Percent_.5 ~ log_metal + log_incinerate + meanPCI + quadPOV, data = i_a_m_b_join)
summary(model9)
plot(model9)
step(model9)

model10 <- lm(formula = Percent_.5 ~ meanPCI, data = i_a_m_b_join)
summary(model10)
plot(model10)

### Regression Visualization #####
par(mfrow=c(2,2), mar = c(2,2,2,2))
plot(model2)
plot(model1)
plot(model3)
plot(model5)
plot(model7)

step(model1) #eliminates variables.... ignore for this model... 

#install.packages("corrplot")
library(corrplot)
df_corr <- i_a_m_b_join %>% 
  select(number_incinerators, number_airports, number_metal_plants, Percent_.5) %>% 
  cor()
df_corr2 <- i_a_m_b_join %>% 
  select(number_incinerators, number_airports, number_metal_plants, meanPCI, quadPOV, Percent_.5) %>% 
  cor()
df_corr2
df_corr # correlation between model's variables. 
par(mfrow=c(1,1))
corrplot(df_corr, method = "ellipse")



#mapping intersections#####
newmap<- ggplot()+
  geom_sf(data = census_sf, color = "Black", fill = "Green", alpha = 0.3, show.legend = "polygon")+
  geom_sf(data = metal_sf, color = 'blue', size = 2, show.legend = "point" )
newmap




```
```{r regression}



```


\newpage
# Summary and Conclusions
\newpage
# References
<add references here if relevant, otherwise delete this section> 
